---
title: "Capstone Project Write-Up"
subtitle: "Fatal Pedestrian and Cyclist Crash Risk Factors in Oregon: A Spatial and Predictive Analysis"
author:
  - "Simon Thompson"
  - "Rohan Srinivasa Babu"
date: today
bibliography: refs.bib
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
  DATA 510 write-up skeleton (Week 5 Quarto lecture).

  Copy this file to your DS3 repo:
    deliverables/M4-writeup-draft/writeup.qmd

  Week 5: paste prose from your M1 proposal into Introduction, Data, Ethics, Methods.
  Week 7+: replace stubs with real results from notebooks/ and src/.
  Week 12 (M4): methods and findings should be stable; render PDF + HTML for submission.
-->

# Title, Team, and Affiliations

**Project title:** *Fatal Pedestrian and Cyclist Crash Risk Factors in Oregon: A Spatial and Predictive Analysis*

| Simon Thompson | Owner Product Lead | sthompson2@willamette.edu |
| Rohan Srinivasa Babu | Owner Team | rsrinivasababu@willamette.edu |


**Peer Stakeholder POs:**
|Aiyana Brown | Peer Stakeholder | acbrown2@willamette.edu |
|Amaya Supancich-McCord | Peer Stakeholder | asupancichmccord@willamette.edu |
|Seira Ramchandani | Peer Stakeholder | sramchandani@willamette.edu |

**Repository:** `https://github.com/S1mT/data510-OregonPedSafety`

# Abstract

*(150 to 250 words after you have results. Summarize: problem, approach, key finding, impact, limitation. Write this section last.)*

# Introduction: Problem and Research Questions

<!-- Paste and revise from M1 proposal Section 2. Note any changes since M1. -->

**Problem statement:**

Pedestrians and cyclists remain among the most vulnerable users of the transportation network. While motor vehicle safety has improved substantially over recent decades, pedestrian and cyclist fatalities continue to represent a significant public health and transportation safety concern. Understanding where these fatalities occur and the environmental, roadway, and demographic factors associated with them can help transportation planners prioritize interventions and allocate limited safety resources more effectively.

This proposal is consistent with the project Vision and Mission established in the project charter. We seek to identify factors associated with fatal pedestrian and cyclist crashes and provide evidence-based recommendations for transportation safety improvements.

**Primary Research Question:**

Which environmental, roadway, and temporal factors are most strongly associated with fatal pedestrian and cyclist crashes in Oregon between 2015 and 2024?

**Secondary Research Question:**

Are fatal pedestrian and cyclist crashes disproportionately concentrated in areas with certain identifiable demographics (race, ethnicity, income, etc.) in Oregon between 2015 and 2024?

**Changes since M1 proposal:** *(none yet, or explain pivot)*

# Impact and Stakeholders

<!-- Paste from M1 proposal Section 3. -->

Pedestrians and Cyclists: Pedestrians and cyclists experience the highest levels of physical vulnerability during roadway collisions. Findings may help identify locations requiring targeted safety improvements.

Oregon Department of Transportation (ODOT): ODOT maintains statewide transportation safety initiatives and may use project findings to better understand risk factors associated with fatal pedestrian and cyclist crashes.

City/County Transportation Planners: Local planners may benefit from spatial analyses identifying high-risk features such as traffic device (STOP sign, YIELD sign, signal), roadway surface, sidewalk presence, etc. for both retroactive and proactive infrastructure investment.

Public Health and Community Organizations: Organizations focused on injury prevention, transportation equity, and community mobility may use findings to support advocacy and planning efforts.


# Related Work and Background

<!-- Brief cited context. Example: -->

Example: Wildfire smoke and respiratory health are well studied at regional scale [@reid2016; @liu2016]. This project applies county-level public data to Oregon specifically.

# Data and Data Engineering

<!-- Paste from M1 Sections 4 and 5; update with actual pipeline state after M2. -->

**Sources:**

| Source | Role | Access |
|--------|------|--------|
| | | |

**Engineering summary:** Raw data in `data/raw/`; processed panel in `data/processed/`; ingest scripts in `src/ingest/`.

```{r}
#| label: tbl-data-qc
#| tbl-cap: "Example QC summary (replace with your panel)"
#| echo: true

# Replace with your project data when ready:
# panel <- readRDS(here::here("data/processed/analysis_panel.rds"))
# nrow(panel)
knitr::kable(data.frame(
  check = c("Rows", "Counties", "Date range"),
  status = c("stub", "stub", "stub")
))
```

# Ethics and Responsible Use

<!-- Paste from M1 Section 8; expand with project-specific limits. -->

- **Privacy:**
- **Fairness / equity:**
- **Limitations of setting:**
- **Non-causal framing:** *(if observational)*

# Methods and Evaluation

<!-- Paste from M1 Section 6; tighten once analysis is fixed. -->

**Design:**

**Primary analysis:**

**Evaluation metrics:**

**Train / validation / holdout:**

```{r}
#| label: fig-methods-placeholder
#| fig-cap: "Replace with a methods diagram or exploratory plot"
#| echo: false
#| fig-width: 5
#| fig-height: 3

plot(1, 1, type = "n", xlab = "", ylab = "", main = "Methods figure placeholder")
text(1, 1, "Replace after M2/M3")
```

# Results and Visualizations

<!-- Fill after M2 data summary and M3 poster figures. -->

*(Stub: key findings with @fig-methods-placeholder and tables.)*

# Discussion, Limitations, and Future Work

**Interpretation:**

**Limitations:** *(tie to actual data and design)*

**Future work:**

# Conclusions and Recommendations

*(Short, stakeholder-facing takeaways.)*

# Reproducibility and References

**Reproduce main results:**

```bash
cd your-repo
quarto render deliverables/M4-writeup-draft/writeup.qmd
```

**Dependencies:** `requirements.txt` or `renv.lock` in repo root.

**Data:** URLs and approval status documented in `docs/data-dictionary.md`.

# References

::: {#refs}
:::

<!-- Create refs.bib alongside this file. Example entries: -->

<!--
@article{liu2016,
  title={Wildfire-specific fine particulate matter and risk of hospital admissions in urban and rural counties},
  author={Liu, J. C. and others},
  journal={Epidemiology},
  year={2016}
}

@article{reid2016,
  title={Critical review of health impacts of wildfire smoke exposure},
  author={Reid, C. E. and others},
  journal={Environmental Health Perspectives},
  year={2016}
}
-->
